---
title: "Causal Inference: Milestone2"
author: "Na Nguyen and Charles Batsaikhan"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r , warning=FALSE, message= FALSE, include = FALSE}
library(dplyr)
library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(readr)
library(dagitty)
library(EValue)
library(MatchIt)
library(broom)
library(marginaleffects)
library(scales)

final <- read.csv('final.csv')

```

# Introduction

In the 1920s, Minneapolis introduced racial covenants to prevent non-white people from occupying and owning land. As Minnesota witnessed the wave of the First Great Migration, it also legitimized housing discrimination via restrictive clauses inserted into property documents since the presence of non-White, but especially Black people, was regarded to decrease property value. This racist real estate practice paved the way to redlining in the 1940s to decrease homeownership rates of African American, contributing to their generational dispossession and displacement. (Source, Source)

While white areas were endorsed with beautification projects and accessible green spaces, Black residents saw their neighborhoods surrounded with highways, factories, and overall exposure to hazard. And the legacy of residential segregation has led to stark environmental injustice that tremendously ails Black, Latino, and Native American families in Minneapolis:

-   6x asthma death risk
    -   Among those who were under age 65, the asthma death rate for Black Minnesotans was 6 times higher than it was for white Minnesotans.
-   3x higher infant mortality
    -   Infant mortality rate for Native Americans is nearly 3x higher than that of white Minnesotans.
-   9x positive COVID-19 tests
    -   Latinx Minnesotans are testing positive at more than 4 times the overall population and 9 times the white population.
-   Over 2x the cancer risk
-   Black Minnesotans have 2.25 times the cancer risk as the average Minnesotan (Source)

There has been extensive research that definitively proves disproportionate pollution linked to geographically disadvantaged and vulnerable communities. However, these research tend to merely provide a snapshot of this picture at a moment in time. Therefore, we would like to contextualize these single portraits of inequality by investigating the relationship between modern day environmental inequality in Minneapolis and 1920’s racial covenants practice.

The research question we would like to explore is as follows: Looking at historically racially covenanted neighborhoods in Minneapolis and present day environmental data, can we still see its impact in the disparity of climate change effects? In other words, *did the introduction of racial covenants in the 1920s cause air pollution inequality in the city of Minneapolis that we see today?*

# Data 

## Sources

## Data Exploration

In this boxplot, we can observe that between the socially vulnerable (Treated) and non-socially vulnerable (Control) groups, the median for air pollution is higher for the treated group by 0.3 tons of PM2.5 emissions.

```{r, echo = FALSE}
median_calc <- final %>% 
  mutate(svi_index = as.character(svi_index)) %>%
  group_by(svi_index) %>%
  summarize(median_air_pollution = median(air_pollution, na.rm = TRUE))

final %>%
  mutate(svi_index = as.character(svi_index)) %>%
  ggplot() +
  geom_boxplot(aes(x=svi_index, y = air_pollution)) +
  geom_text(data = median_calc, aes(x = svi_index, y = median_air_pollution, 
                                    label = paste(round(median_air_pollution, 2))),
            vjust = -1, color = "black", size = 5) +
  theme_minimal()
```


# Methods

Moving forward, we will work with the following variables at the census tract level in the city of Minneapolis:

-   `has_racial_covenants`: whether or not a census tract previously had a racial covenant on its property
-   `holc_grade`: the census tract's Home Owner Loan Corporation (HOLC) neighborhood grades: A, B, C, or D (corresponding to A = “Best”; B = “Still Desirable”; C = “Declining”; and D = “Hazardous” designations, respectively)
-   `svi_index`: whether or not a census tract is considered as socially vulnerable by the CDC
-   `tree_canopy`: the percentage of tree canopy area of a census tract.
-   `air_pollution`: the emission (tons) of PM2.5 from permitted facilities/factories in a census tract.

**Note: A handful of census tracts do not have a `holc_grade` because they reside in downtown business areas of Minneapolis, not residential areas**

## Approach

-   We conducted data cleaning and transformation since we started out with mostly spatial data.
-   Then, we conducted matching on `holc_grade` in order to control for confounders using 5 matching methods for an ATE estimand: Nearest Neighbors (k=2), Full Matching, Exact Matching, Coarsened Exact Matching, and Subclass Matching.
-   Then, we build a regression model with `svi_index` as our treatment variable (0 if a census is not socially vulnerable ), `air_pollution` as our outcome variable (tons of PM2.5 emissions).

## Our Causal DAG

According to our DAG, our control set includes the following variables: `holc_grade`, and `tree_canopy`

```{r, warning=FALSE, message= FALSE, echo=FALSE}
dag<- dagitty(
  'dag {
bb="0,0,1,1"
"infrastructure divestment (U)" [latent,pos="0.408,0.406"]
"urban/rural" [latent,pos="0.181,0.096"]
air_pollution [outcome,pos="0.693,0.604"]
has_racial_covenant [pos="0.180,0.246"]
holc_grade [pos="0.182,0.399"]
svi_index [exposure,pos="0.282,0.603"]
tree_canopy [pos="0.651,0.403"]
"infrastructure divestment (U)" -> tree_canopy
"infrastructure divestment (U)" -> svi_index
"infrastructure divestment (U)" -> air_pollution
"urban/rural" -> has_racial_covenant
has_racial_covenant -> holc_grade
holc_grade -> "infrastructure divestment (U)"
holc_grade -> svi_index
svi_index -> air_pollution
tree_canopy -> air_pollution
}

')

plot(dag)
```

## Matching

We originally have 116 total observations, 19 untreated and 97 treated census tracts. In order to control for our confounder---`holc_grade`, we conducted 5 matching methods for an ATE estimand: Nearest Neighbors (k=2), Full Matching, Exact Matching, Coarsened Exact Matching, and Subclass Matching. Our criteria for choosing a final matching method are:

-   Standard Mean Difference for all `holc_grade`'s are well within the 0-0.1 threshold.
-   Yielding a large enough matched sample size.

In the end, we opted for the Exact Matching Method. As opposed to other methods that resulted in rather large Standard Mean Differences ( \>= 0.4), Exact Matching will keep our SMD at the minimum (0). However, because there are a handful of census tracts that do **not** have a `holc_grade`, our final matched sample size for treated is smaller: 19 untreated and 72 treated. The limited sample size may affect the generalizability of the results, so the findings should be interpreted with caution, and future studies with larger sample sizes would be needed to validate the conclusions.

Nonetheless, this method is the most optimal as it helps us to be confident in our later estimates for the causal relationship between our outcome and treatment variables.

## Estimate Results

1.  Estimate (0.967): The coefficient for `svi_index` is 0.967. This suggests that of all census tracts in Minneapolis that are matched on, the air quality is worse for socially vulnerable neighborhoods, with the average annual PM2.5 emissions increasing by 0.967 tons, holding `tree_canopy_area` constant.

2.  Std. Error (0.298): The standard error of the coefficient is 0.298, which reflects the variability in the estimate of the coefficient. This is a relatively small st.Error.

3.  Pr(\>\|z\|) (0.001): The p-value is 0.001, which is much smaller than the commonly used significance threshold of 0.05. This indicates that the relationship between `svi_index` and `air_pollution` is statistically significant.

4.  95% Confidence Interval (0.889, 2.06): Since the interval does not include zero and it's quite narrow, this suggests that there is evidence of a non-zero relationship between `svi_index` and `air_pollution.` We are 95% confident that this CI (0.889, 2.06) represents a range of plausible values for this effect of `svi_index` on `air_pollution`.

```{r, echo = FALSE}

match_out_exact <- matchit(
    svi_index ~ holc_grade,
    data = final,
    method = "exact",
    distance = NULL,
    estimand = "ATE"
)

match_out_exact_summ <- summary(match_out_exact, interactions = TRUE)

match_out_exact <- match.data(match_out_exact)

mod <- lm(air_pollution ~ svi_index*tree_canopy_area, data = match_out_exact, weights = weights)

avg_comparisons(
    mod,
    variables = "svi_index",
    vcov = ~subclass,
    newdata = match_out_exact
)

```

## Sensitivity Analysis


```{r, echo = FALSE}
evalues.OLS(est= 0.967, se=0.448, sd=sd(match_out_exact$air_pollution))

```

If the estimate of the causal effect is 0.967, social vulnerability would have to be 2.76 times more common in census tracts with infrastructure divestment than those that don't.

If the estimate of the causal effect is 0.967, infrastructure divestment would have to be 2.76 times more common in census tracts with high pollution areas than those that don't.

# Final Conclusion

# Code

Full Matching

```{r, eval = FALSE}
#| code-fold: true
match_out_full <- matchit(
    svi_index ~ holc_grade,
    data = final,
    method = "full",
    distance = "glm",
    estimand = "ATE"
)

match_out_full_summ <- summary(match_out_full, interactions = TRUE)

plot(match_out_full_summ)
```

Subclass matching

```{r, eval = FALSE}
#| code-fold: true
match_out_subclass <- matchit(
    svi_index ~ holc_grade,
    data = final,
    method = "subclass",
    subclass = 5,
    distance = "glm",
    estimand = "ATT"
)

# Compute balance statistics overall across all subclasses
match_out_subclass_summ <- summary(match_out_subclass, interactions = TRUE)
plot(match_out_subclass_summ)
```

Coarsened Exact Matching

```{r,  eval = FALSE}
#| code-fold: true
match_out_cem <- matchit(
    svi_index ~ holc_grade,
    data = final,
    method = "cem",
    distance = NULL,
    estimand = "ATE"
)

match_out_cem_summ <- summary(match_out_exact, interactions = TRUE, un = FALSE)
match_out_cem_summ
```

Exact Matching

```{r,  eval = FALSE}
#| code-fold: true
match_out_exact <- matchit(
    svi_index ~ holc_grade,
    data = final,
    method = "exact",
    distance = NULL,
    estimand = "ATE"
)

match_out_exact_summ <- summary(match_out_exact, interactions = TRUE, un = FALSE)
match_out_exact_summ

```

Estimates

```{r,  eval = FALSE}
#| code-fold: true

mod <- lm(air_pollution ~ svi_index*tree_canopy_area, data = match_out_exact, weights = weights)

avg_comparisons(
    mod,
    variables = "svi_index",
    vcov = ~subclass,
    newdata = match_out_exact
)

```

Evalues

```{r,  eval = FALSE}
#| code-fold: true

evalues.OLS(est= 0.967, se=0.448, sd=sd(match_out_exact$air_pollution))

```
